\documentclass{article}
\usepackage{lettrine}
\usepackage{epigraph, varwidth}

\begin{document}
\renewcommand{\epigraphsize}{\small}
\setlength{\epigraphwidth}{1.0\textwidth}

\section {Authority}

\lettrine[lines=2]{G}{overnment, the legal system,} school, organized religions, the media, corporations, various economic concerns \dots there are a number of powerful, pervasive institutions that want to control you. It's either their overt purpose or a byproduct of their very existence. In order not to disintegrate, they must have at least a piece of you, if not all of you, for at least part of your life, if not all of it. They might not have been created for nefarious purposes, but this is nonetheless the effect they often have. They flatten and deaden. They exploit and punish. They are never happy with the amount of power they have; they always seek to expand the breadth of their reach and the strength of their grip. This has not gone unnoticed by many of the finest and bravest minds throughout history\footnote{The \emph{emphases} in quotations are mine.} \dots

\epigraph
{
    No man has any natural authority over his fellow men.
}
{
    ---\textit{Jean-Jacques Rousseau}
}
\epigraph
{
    Reexamine all you have been told at school or church or in any book, dismiss whatever insults your own soul.
}
{
    ---\textit{Walt Whitman}
}
\epigraph
{
    I must Create a System,\\
    or be enslav'd by another Man's.
}
{
    ---\textit{William Blake}
}
\epigraph
{
    You never change things by fighting the existing reality. To change something, build a new model that makes the existing model obsolete.
}
{
    ---\textit{R. Buckminster Fuller}
}
\epigraph
{
    Do not wait for leaders; do it alone, person to person.
}
{
    ---\textit{Mother Teresa}
}
\epigraph
{
    The whole history of the progress of human liberty shows that all concessions yet made to her august claims have been born of earnest struggle.\dots \emph{If there is no struggle, there is no progress.} Those who profess to favor freedom, and yet depreciate agitation, are men who want crops without plowing up the ground. They want rain without thunder and lightning. They want the ocean without the awful roar of its many waters. This \emph{struggle} may be a moral one; or it may be a physical one; or it may be both moral and physical; but must be a struggle. Power concedes nothing without a demand. It never did and it never will. Find out just what any people will submit to, and you have found out the exact amount of injustice and wrong which will be imposed upon them; and these will continue till they are resisted with either words or blows, or [with] both. \emph{The limits of tyrants are prescribed by the endurance of those whom they oppress.}
}
{
    ---\textit{Frederick Douglass}
}
\end{document}
